Prosper Loan Data Analysis by Jason Yang
========================================================

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, comment=NA)

```

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(plyr)
library(dplyr)
library(gridExtra)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
prosperLoanData <- read.csv(file='./prosperLoanData.csv', header=TRUE, sep=",")

```


In this analysis, we explore loan data from Prosper. This data set contains 
113,937 loans with 81 variables on each loan, including loan amount, borrower 
rate (or interest rate), current loan status, borrower income, borrower 
employment status, borrower credit history, and the latest payment information.

# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots_structure}
str(prosperLoanData)
```
Our dataset consists of 81 variables and about 114,000 observations.


```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots1}
# summary(prosperLoanData);

ggplot(aes(x=IncomeRange), data = prosperLoanData) + 
  geom_histogram(stat = 'count') +
  theme(axis.text.x = element_text(angle=90, hjust = 1)) +
  scale_x_discrete(limits=c('$0','$1-24,999','$25,000-49,999', '$50,000-74,999',
                            '$75,000-99,999','$100,000+','Not displayed',
                            'Not employed'))
```

To start the investigation, I wanted to explore what type of customers Prosper
make loans to. By plotting the income range, we can see that most of the 
borrowers are between income ranges of $25,000 and $75,000. I wonder what this 
plot looks like across other categorical variables of employment status, 
homeownership, and loan status.


```{r}
levels(prosperLoanData$LoanStatus) <- c('Cancelled','Chargedoff', 'Completed', 
                                        'Current','Defaulted',
                                        'FinalPaymentInProgress','Past Due',  
                                        'Past Due', 'Past Due','Past Due', 
                                        'Past Due', 'Past Due')

levels(prosperLoanData$LoanStatus)
```

Combined all Past due statuses into a single past due status.


```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots2}
ggplot(aes(x=EmploymentStatus), data = prosperLoanData) + 
  geom_histogram(stat = 'count') +
  theme(axis.text.x = element_text(angle=90, hjust = 1))

ggplot(aes(x=IsBorrowerHomeowner), data = prosperLoanData) + 
  geom_histogram(stat = 'count')

ggplot(aes(x=LoanStatus), data = prosperLoanData) + 
  geom_histogram(stat = 'count') +
  theme(axis.text.x = element_text(angle=90, hjust = 1)) +
  scale_x_discrete(limits=c('Cancelled','Chargedoff', 'Completed', 'Current',
                            'Defaulted', 'FinalPaymentInProgress', 'Past Due'))
```

Most borrowers are employed, this categorical variable has 2,255 null values. 
The number of borrowers that is a homeowner is about the same as the number not
owning a home. A majority of the loan that was taken out has a status of 
completed or current.


```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots3}
ggplot(aes(x=BorrowerAPR), data = prosperLoanData) + 
  geom_histogram() 

prosperLoanData$CreditScoreAverage <- (prosperLoanData$CreditScoreRangeLower +
                                      prosperLoanData$CreditScoreRangeUpper)/2

ggplot(aes(x=CreditScoreAverage), data = prosperLoanData) + 
  geom_histogram(binwidth = 10) +
  coord_cartesian(xlim=c(450,850)) +
  xlab("Average Credit Score") +
  scale_x_continuous(breaks = seq(450,850,50))

ggplot(aes(x=LoanOriginalAmount), data = prosperLoanData) +
  geom_histogram(binwidth = 500) +
  scale_x_continuous(breaks = seq(0,35000,5000))

```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots4}
summary(prosperLoanData$BorrowerAPR)
summary(prosperLoanData$CreditScoreAverage)
summary(prosperLoanData$LoanOriginalAmount)

```

The lowest APR was 0.00653 and the highest APR was 0.512. On average, a 
borrower had an APR of 0.219, which was higher than the median APR of 0.210. 
Visually, we can see the Borrower APR graph skewed slightly to the right, with 
a two local maximum peaks to the right of 0.25. I wonder how higher APR is 
connected to homeownership and employment status. \
To analyze credit score, I introduced a new variable called 
"CreditScoreAverage." This new variable is the average of the upper and lower 
credit score in our dataset. I've trimmed the graph to be between 450 and 850, 
which was where most of the data was. Most users had a credit score of around 
690. \
Viewing the original amount of loan incurred, common amounts include increments 
of thousands (i.e. $1,000; $2,000; etc.) in low ranges and increments of 5 
thousands up to $35,000. The median loan original amount was $6,500.

# Univariate Analysis

### What is the structure of your dataset?
Our dataset consists of 81 variables and about 114,000 observations. Borrowers 
are categorized into income ranges, homeownership, and loan status. \
Observations: \
- The number of homeowners and non-homeowners are about the same  \
- Median credit score is 690 \
- Median and mean APR are 0.210 and 0.219, respectively \
- Most loan status are complete or current. \
- A majority of borrowers make loans below $10,000.


### What is/are the main feature(s) of interest in your dataset?
For me, the main features of interest are borrower's credit score, original 
loan amount, and borrower's APR. I suspect borrower's APR can be modeled using 
borrower's credit score in some combination of other variables. I think APR is 
heavily dependent on credit score and loan amount, but other categorical 
variables may reveal additional relationships.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
Other features of interest include homeownership, employment status, and 
income range. It would be interesting to investigate the dataset using loan 
status feature to see which loans are deliquent (charged off, defaulted, past 
due).

### Did you create any new variables from existing variables in the dataset?
For the univariant analysis, I combined all Past due statuses into a single 
past due status. I also created the "CreditScoreAverage" variable. 


# Bivariate Plots Section

```{r echo=FALSE, Bivariate_Plots}
ggplot(aes(x=CreditScoreAverage, y =BorrowerAPR), data = prosperLoanData) +
  geom_point(alpha = 0.01,
             position = position_jitter(h = 0),
             color = 'blue')
```

As one would expect, higher credit scores are usually related with lower APR. 
However, credit scores do not seem to be the only deciding factor when 
determining APR. From the graph, there are distinct horizontal lines at 0.3 and 
0.375 APR. I think plotting the borrower APR and loan amount graph will help.


```{r echo=FALSE, Bivariate_Plots2}
ggplot(aes(x=LoanOriginalAmount, y =BorrowerAPR), data = prosperLoanData) +
  geom_point(alpha = 0.01,
             position = position_jitter(h = 0),
             color = 'blue')
```

There is a weak correlation between loan amount and borrower APR. It seems 
like when borrowing above $30,000, APR is more likely to be between 0.2 and 0.1,
but I think this has more to do with the type of users borrowing in high 
numbers. Regardless, there is a weak correlation in that APR goes down when 
borrowing at higher amounts, which makes sense. Mortgage interest rates are 
usually lower than credit card interest rates.


```{r echo=FALSE, Bivariate_Plots3}
ggplot(aes(x=IncomeRange, y =CreditScoreAverage), data = prosperLoanData) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle=90, hjust = 1)) +
  scale_x_discrete(limits=c('$0','$1-24,999','$25,000-49,999', '$50,000-74,999',
                            '$75,000-99,999','$100,000+','Not displayed',
                            'Not employed')) +
  coord_cartesian(ylim=c(300,850)) +
  scale_y_continuous(breaks = seq(300,850,50))

by(prosperLoanData$CreditScoreAverage, prosperLoanData$IncomeRange, summary)

```

Credit score had less dependency on income range than I expected. All income 
ranges except "Not displayed" had a median credit score average of 689.5 or 
709.5. 

```{r echo=FALSE, Bivariate_Plots4}
ggplot(aes(x=IncomeRange, y =BorrowerAPR), data = prosperLoanData) +
  geom_point(alpha = 0.01,
             position = position_jitter(h = 0),
             color = 'blue') +
  theme(axis.text.x = element_text(angle=90, hjust = 1)) +
  scale_x_discrete(limits=c('$0','$1-24,999','$25,000-49,999', '$50,000-74,999',
                            '$75,000-99,999','$100,000+','Not displayed',
                            'Not employed'))

```

I decided to plot the IncomeRange vs BorrowerAPR plot as a point plot instead 
of a boxplot to visualize where most loans are made. There's a distinct line of 
loans made with 0.35 APR over income ranges of $1 and $100,000+. Let's 
visualize what type of loans people within an income range is making.

```{r echo=FALSE, Bivariate_Plots5}
ggplot(aes(x=IncomeRange, y =LoanOriginalAmount), data = prosperLoanData) +
  geom_point(alpha = 0.01,
             position = position_jitter(h = 0),
             color = 'blue') +
  theme(axis.text.x = element_text(angle=90, hjust = 1)) +
  scale_x_discrete(limits=c('$0','$1-24,999','$25,000-49,999', '$50,000-74,999',
                            '$75,000-99,999','$100,000+','Not displayed',
                            'Not employed'))

```

Most loans were made by borrowers in the $25,000 and $75,000 range. Users with 
higher income are more likely to make bigger loan amounts.


```{r echo=FALSE, Bivariate_Plots6}
ggplot(prosperLoanData, aes(CreditScoreAverage, fill=IsBorrowerHomeowner)) +
         geom_histogram()

by(prosperLoanData$CreditScoreAverage, prosperLoanData$IsBorrowerHomeowner, summary)
```

Borrowers who are homeowners have slightly higher credit scores (median=709.5)
over borrowers who do not own homes (median=689.5). Although the medians are 
very similar, the two group's mean differs much more. Borrowers who are 
homeowners have a mean credit score of 714.3, and borrowers who are not 
homeowners have a mean score of 675.3. I think we'll be able to visually see 
this  when we plot credit score vs. APR with homeownership context in the 
multi-variant section.

```{r echo=FALSE, Bivariate_Plots7}
ggplot(aes(x=EmploymentStatus, y =CreditScoreAverage), data = prosperLoanData) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle=90, hjust = 1)) +
  coord_cartesian(ylim=c(300,850)) +
  scale_y_continuous(breaks = seq(300,850,50))

by(prosperLoanData$CreditScoreAverage, prosperLoanData$EmploymentStatus, summary)

```

It looks like credit scores vary much more on employment status than on income 
range. I'm surprised to see that the credit scores of employed and not employed
users are about the same at mean=708.8, 703.0 and above that of full-time status
(mean=678.9)

```{r echo=FALSE, Bivariate_Plots8}
ggplot(prosperLoanData, aes(x=LoanStatus, y =BorrowerAPR)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle=90, hjust = 1))

by(prosperLoanData$BorrowerAPR, prosperLoanData$LoanStatus, summary)

```

A borrower's APR is determined before the loan is made. It's interesting to 
see that the bad loan statuses (charged off, defaulted, past due) have higher 
APR than that of good loan statuses (completed, current, final payment). I 
wonder if higher APR was knowingly assigned to borrowers who will have trouble 
paying back, or is it because of higher APR that resulted in bad loan statuses.

```{r echo=FALSE, Bivariate_Plots9}
ggplot(prosperLoanData, aes(x=LoanStatus, y =CreditScoreAverage)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle=90, hjust = 1))

by(prosperLoanData$CreditScoreAverage, prosperLoanData$LoanStatus, summary)

```

As expected from our previous observation of LoanStatus vs BorrowerAPR, bad 
loan statuses have lower credit score than good loan statuses. The exception is 
the canceled loan status, which has the worst credit score.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?
The relationship between credit score and borrower APR was as expected. Higher 
credit scores yielded lower APR. The relationship between loan amount and 
borrower APR was weaker than I expected. For some reason, I was expecting that 
larger loans would mean higher APR, but it turned out to be the reverse. Larger 
loan amount correlated with lower APR. Which makes sense when I think about it. 
A house mortgage usually has lower interest rate than credit card interest rate.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
The most interesting relationship observed was bad loan statuses are mostly 
associated with high APR. I wonder if higher APR was knowingly assigned to 
borrowers who will have trouble paying back, or is it because of higher APR that
resulted in bad loan statuses. Borrowers with high APR mostly have lower credit 
scores, so in effect, loan issuers know that the borrower may have trouble 
paying the loan back and issued loans with higher APR. However, I wonder how 
much of bad loan status was the result of bad credit score or the fact that 
those borrowers had to pay more due to higher APR. 

### What was the strongest relationship you found?
The strongest relationship was between credit score and borrower APR.

# Multivariate Plots Section


```{r echo=FALSE, Multivariate_Plots}

ggplot(prosperLoanData, aes(x=CreditScoreAverage, y=BorrowerAPR, color=IsBorrowerHomeowner)) +
  geom_point(alpha = 0.05,
             position = position_jitter(h = 0))

```

The legend came out faint. True is cyan and False is salmon. Homeowners have 
higher credit scores and lower APR.

```{r echo=FALSE, Multivariate_Plots2}

ggplot(prosperLoanData, aes(x=LoanOriginalAmount, y=BorrowerAPR, color=IsBorrowerHomeowner)) +
  geom_point(alpha = 0.05,
             position = position_jitter(h = 0))

```

Now this graph is interesting. I expected the loan amount and APR to be fairly 
even between homeowners and non-homeowners up to $20,000. It turns out small 
loans are mostly made by non-homeowners and large loans are made by homeowners. 
The lowest rates are largely given to homeowners.

```{r echo=FALSE, Multivariate_Plots3}

ggplot(prosperLoanData, aes(x=LoanStatus, y =BorrowerAPR, 
                            color=IsBorrowerHomeowner)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle=90, hjust = 1))

```

As expected, homeowners have lower overall APR rate across all loan statuses.

```{r echo=FALSE, Multivariate_Plots4}

ggplot(prosperLoanData, aes(x=IncomeRange, y =CreditScoreAverage, 
                            color=IsBorrowerHomeowner)) +
  geom_point(alpha = 0.05,
             position = position_jitter(h = 0)) +
  theme(axis.text.x = element_text(angle=90, hjust = 1)) +
  scale_x_discrete(limits=c('$0','$1-24,999','$25,000-49,999', '$50,000-74,999',
                            '$75,000-99,999','$100,000+','Not displayed',
                            'Not employed'))

```

Another view of IncomeRange vs. credit score. Homeowners have higher credit 
score and have higher income.

```{r echo=FALSE, Multivariate_Plots5}

ggplot(prosperLoanData, aes(x=CreditScoreAverage, y =BorrowerAPR)) +
  geom_point(aes(color=LoanOriginalAmount), alpha = 0.05,
             position = position_jitter(h = 0)) +
  scale_color_gradient(low = "red", high = "blue")

```

Large loans are mostly made by users with higher credit score, which yields a 
lower APR. Larger loans also reduces APR.

```{r echo=FALSE, Multivariate_Plots6}

ggplot(prosperLoanData, aes(x=CreditScoreAverage, y =BorrowerAPR)) +
  geom_point(aes(color=ProsperScore), alpha = 0.05,
             position = position_jitter(h = 0)) +
  scale_color_gradient(low = "red", high = "blue")

```

As an aside, I wanted to see how prosper scores their borrowers. As expected, 
low APR are given to borrowers with high prosper score. Having a high credit 
score could help in getting low APR, but there seems to be  other variables at 
play when determing prosper score. The gray parts of the graph refer to data 
points that did not have a prosper score. Not all transactions had prosper 
scores tied to them. From the graph it looks like borrowers below a credit score 
of 600 do not have a prosper score. Maybe if your credit score is below a 
certain threshold, other variables take precidence in determining APR over 
prosper score.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

The "IsBorrowerHomeowner" feature was especially helpful in providing 
additional context to the plots. From the bivariant analysis, we determined 
homeowners to have higher credit scores, take higher loans, and have lower APR. 
With the addition of "IsBorrowerHomeowner", it is easier to visualize the 
characteristics of a homeowner.


### Were there any interesting or surprising interactions between features?

I found it interesting that the lowest APR was given to homeowners regardless 
of amount. I expected loan amounts below $5,000 will have about the same APR for
homeowners and non-homeowners.



------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
# Source: https://www.r-graph-gallery.com/264-control-ggplot2-boxplot-colors/
# prosperLoanData$LoanStatusType=factor(ifelse(prosperLoanData$LoanStatus== "", 
# c("Chargedoff","Defaulted","Past Due"), c('Cancelled','Completed','Current', 
# 'FinalPaymentInProgress' )))


ggplot(prosperLoanData, aes(x=LoanStatus, y =BorrowerAPR)) +
  geom_boxplot() +
  ggtitle('Loan Status vs Borrower APR') +
  xlab('Loan Status') +
  ylab('Brrower APR') +
  theme(axis.text.x = element_text(angle=90, hjust = 1),
        plot.title = element_text(hjust = 0.5))


```

### Description One

Bad loan statuses are correlated with higher APR. From this graph, we see bad 
loan statuses (Chargedoff, Defaulted, Past Due) have higher APR than good loan 
statuses (Completed, Current, FinalPaymentInProgress). Cancelled is a neutral 
status because the borrower decided not to take out the loan.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(prosperLoanData, aes(x=CreditScoreAverage, y=BorrowerAPR, 
                            color=IsBorrowerHomeowner)) +
  geom_point(alpha = 0.05,
             position = position_jitter(h = 0)) +
  ggtitle('Average Credit Score vs Borrower APR') +
  xlab('Average Credit Score') +
  ylab('Brrower APR') +
  theme(plot.title = element_text(hjust = 0.5))
```

### Description Two

Having a high credit score and homeownership is correlated with lower APR. A 
majority of users with low credit score does not own a home.

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(prosperLoanData, aes(x=LoanOriginalAmount, y=BorrowerAPR, 
                            color=IsBorrowerHomeowner)) +
  geom_point(alpha = 0.05,
             position = position_jitter(h = 0)) +
  ggtitle('Original Loan Amount vs Borrower APR') +
  xlab('Original Loan Amount ($)') +
  ylab('Brrower APR') +
  theme(plot.title = element_text(hjust = 0.5))
```

### Description Three

Small loans are mostly made by non-homeowners and large loans are mostly made 
by homeowners. The lowest rates are largely given to homeowners regardless of 
loan amount. Higher APR (0.35) are more common with loans below $10,000.

------

# Reflection

Prosper Loan Data set is a very detailed dataset with many features to 
analyze. Not from a finance background, I had to read up on the terminology for 
quite a few words and descriptions for concepts. Something that should be 
obvious like lower APR in larger loan amounts was not immediately obvious to me. 
I think that's one of the challenges in being a data analyst in that we have to 
become familiar with the data before we can ask the right questions. \
In my analysis, I've only explored a small set of features from this dataset. 
There's much more to analyze with this dataset. A future work would be to
explore Prosper's own ratings of borrowers. In the multi-variant section, I 
explored the prosper score feature. We saw borrowers with credit scores below 
600 was not given a Prosper rating. I'm curious to see what other methodologies 
Prosper uses to determine a borrower's APR.